---
title: "Research Unemployment to GDP"
author: "Studentnames and studentnumbers here"
date: "`r Sys.Date()`"
output: pdf_document
---

# Set-up your environment

```{r package_install, include=FALSE}
#install.packages("tidyverse")
#install.packages("readxl")
#install.packages("rworldmap")
#install.packages("dplyr") 
#install.packages("ggplot2")

```

```{r packages}
require(tidyverse)
require(readxl)
require(rworldmap)
require(dplyr)
require(ggplot2)

```

# Unemployment in Europe

Chris van Weerdenburg, Arsema Tewelde, Alan Peralta Camacho, Benjamin Mutlu, Cihan Öztürk, Israa Shidai, Jeremey Li

Tutorial group number: 3

Tutorial lecturer: Jack Fitzgerald

# Part 1 - Identify a Social Problem

## 1.1 Describe the Social Problem

Include the following:

-   Why is this relevant?

-   ....

**Unemployment in the European Union (EU)** is a significant social issue that affects millions of individuals, families, and communities across all countries within Europe. It refers to the condition where people who are willing and able to work are unable to find suitable employment. This issue impacts not only those directly affected by this dilemma but also the broader economy and social cohesion within Europe. This is relevant because:

High unemployment rates can lead to decreased consumer spending, lower tax revenues, and increased government spending on social welfare programs. This slows down economic growth and can also hinder post-crisis recovery efforts, for example COVID-19.

It also causes long-term unemployment, which can lead to poverty, social isolation, and reduced access to healthcare and education. It often disproportionately affects young people, older workers, migrants, and those with lower educational qualifications, contributing to widening social inequality.

Youth unemployment is particularly pressing in many EU countries, with some regions experiencing rates above 30%. This delays independence, contributes to brain drain (migration of skilled youth), and weakens future labor market prospects.

Unemployment is closely linked to poor mental health outcomes such as depression, anxiety, and a loss of purpose or identity. The longer someone remains unemployed, the more profound the psychological effects. Many studies show that unemployment leads to depressive symptoms but also a lot of stress, because as a person you lack stability and that causes a feeling of fear.

Here are two important sources that highlight the serious consequences of unemployment. The first source comes from an article by the U.S. Department of Health called *Healthy People 2030*. This report states that work is one of the five social factors of health. According to this report, unemployment can actually cause physical conditions such as high blood pressure, but also psychological problems (Employment – Healthy People 2030 \| odphp.health.gov, z.d.).

Another source that highlights the social effects of unemployment is a report from *ScienceDirect*, written by Pohlan. This report states that losing your job in the long term leads to a decrease in social integration, which means actively participating in social activities, and that leads to long-lasting depression (Pohlan, 2019).

Persistent unemployment can erode trust in institutions, fuel populism and extremism, and lead to social unrest. Citizens may feel disillusioned with governments and the EU itself, challenging the union’s stability and unity.

There have been quite a few studies on the social problem of unemployment, but there is no research that deeply analyzes the relationship between unemployment and economic growth such as GDP. This report focuses on the economic side by looking at how unemployment affects the economy, as well as the social consequences it brings.

# Part 2 - Data Sourcing

We have the following datasets:

-   **Unemployment by sex, age and duration of unemployment (Eurostat)**

-   **Real GDP growth rate - volume (Eurostat)**

-   **Population, aged 15 - 74 - EU labour force survey (Eurostat)**

These datasets are published and sourced by Eurostat. Eurostat is the statistical office of the European Union. They source their data through the ESS which is the European Statistical System. These datasets are also collected by each country's own National Statistical Institute for example the data Eurostat collects from the Netherlands is from the CBS (Centraal Bureau Statistiek). This data collection is done according to established EU regulations and standards, ensuring a degree of harmonization across the EU. The national authorities then transmit the verified and analyzed data to Eurostat.

These datasets have some differences but also similarities.

**Key differences:**

-   GDP per capita relates to overall economic performances, while youth unemployment reflects specific labor market outcomes and the labour force refers to the amount of people between age 15 and 74 who are either employed or unemployed.

-   GDP includes the whole population; youth unemployment only concerns young people who are economically active and the labour force includes people between the ages 15 and 74.

-   GDP is measured in economic value (adjusted for PPS); youth unemployment is a rate (percentage); the labour force is measured in amount of people.

**Complementary:**

-   A country may have high GDP per capita but also high youth unemployment, indicating that economic gains are not evenly distributed or accessible.

-   Using our datasets together provides a more comprehensive view of socio-economic well-being, economic performance (GDP) and inclusiveness or opportunity (youth employment).

-   This combination can be especially useful for analyzing whether economic growth is benefiting younger generations.

-   Usage of these different datasets gives us a broad insight and amount of data that we could use each and every aspect of the data that we sourced, in order to create clear and explicit plots.

These two datasets were a perfect match for our topic because they gave us an in depth look into the correlation between the GDP growth rate and Unemployment rates but also the unemployment rates and sex, age and duration of unemployment these datasets were crucial for us to conclude whether Covid-19 had an impact on unemployment rates. We added in another dataset which provided the data for the total population aged from 15 to 74 usually residing in Member States in private households, except for persons living in collective or institutional households (e.g. hospitals, residential homes, prisons, military barracks, religious institutions, etc.). With use of this data we could calculate the ratio of population and amount of unemployed. Since with the datasets we had used it did not provide this information. That made it quite hard to properly understand how many people were unemployed since some countries have much bigger populations than others.

[Metadata](https://ec.europa.eu/eurostat/cache/metadata/en/lfsa_esms.htm)Dataset informationAdd to ‘My datasets’

## Selection

The datasets we used were quite complete however some countries the data was not sourced for example: Bosnia and Herzegovina and Lichtenstein. However we did add in a new data set because the other data sets did not give as a good insight about the percentage of the population being unemployed. It just gave us the amount of people that are / were unemployed. This was a slight hiccup since for example Germany has a bigger population than Norway. Other than that were there no complaints about the datasets, and were they very explicit.

## 2.1 Load in the data

```{r loading_data, include=FALSE}
GDP_df <- read_xlsx(path = "tec00115_page_spreadsheet.xlsx", "Sheet 1") 
Unemploymentlang <- read_xlsx(path = "lfsa_ugad$defaultview_spreadsheet.xlsx", "Sheet 1")
Total_Work_population <- read_xlsx(path = "tipslm16_page_spreadsheet.xlsx", "Sheet 1") 
```

midwest is an example dataset included in the tidyverse package

## 2.2 Provide a short summary of the dataset(s)

```{r echo=TRUE}
head(GDP_df[,(1:3)], n = 5)
head(Unemploymentlang[,(1:3)], n = 5)
head(Total_Work_population[,(1:3)], n = 5)
```

In this case we see 28 variables, but we miss some information on what units they are in. We also don't know anything about the year/moment in which this data has been captured.

These are things that are usually included in the metadata of the dataset. For your project, you need to provide us with the information from your metadata that we need to understand your dataset of choice.

## 2.3 Describe the type of variables included

Think of things like:

-   Do the variables contain health information or SES information?

-   Have they been measured by interviewing individuals or is the data coming from administrative sources?

*For the sake of this example, I will continue with the assignment...*

# Part 3 - Quantifying

## 3.1 Data cleaning

We will start with the unemployment dataset. We will remove the first two columns, as they are not relevant for our analysis.

```{r data_cleaning_Unemployment, echo=FALSE}

Unemploymentlang <- Unemploymentlang[14:49, ]
Unemploymentlang <- Unemploymentlang[, -c( 3, 5, 7 ,9 ,11 ,13 ,15 ,17 ,19 ,21)]
rownames(Unemploymentlang) <- NULL
colnames(Unemploymentlang)[2:11] <- as.character(2015:2024)
colnames(Unemploymentlang)[1]  <- "Country"

```

```{r show_data, echo=TRUE}
head(Unemploymentlang, n = 3)
```

```{r data_cleaning_GDP, include=FALSE}
GDP_df <- GDP_df[10:52, ]
rownames(GDP_df) <- NULL
GDP_df <- GDP_df[, colSums(!is.na(GDP_df)) > 0]
GDP_df <- GDP_df[, -c( 5, 12, 14, 16, 18)]
colnames(GDP_df)[2:13] <- as.character(2013:2024)
GDP_df[ , as.character(2013:2024)] <- lapply(GDP_df[ , as.character(2013:2024)], as.numeric)
GDP_df <- GDP_df[-c(1,2,3),]
colnames(GDP_df)[1]  <- "Country"

```

```{r show_data_GDP, echo=TRUE}
head(GDP_df, n = 3)

```

```{r data_cleaning_Total_work_pop, include=FALSE}
Total_Work_population <- Total_Work_population[13:39, ]
Total_Work_population <- Total_Work_population[, -c(2:41)]
Total_Work_population <- Total_Work_population[, -c( 3, 5, 7 ,9 ,11 ,13 ,15 ,17 ,19 ,21)]
rownames(Total_Work_population) <- NULL
colnames(Total_Work_population)[2:11] <- as.character(2015:2024)
Total_Work_population[, 2:11] <- lapply(Total_Work_population[, 2:11], function(x) as.numeric(as.character(x)))
colnames(Total_Work_population)[1]  <- "Country"

```

```{r show data unemp_men}
head(Total_Work_population, n = 3)
```

## 3.2 Generate necessary variables

Variable 1: Classification by Economic Growth

To compare unemployment rates across Europe, we divided the countries into three equally sized groups based on total GDP growth. This allows us to compare unemployment rates across different levels of economic growth.

```{r gen_var1}

GDP_df <- GDP_df %>%
  rowwise() %>%
  mutate(
    GDP_growth_total = round((prod(1 + c_across(`2015`:`2024`) / 100, na.rm = TRUE) - 1) * 100, 2)
  ) %>%
  ungroup()

merged_df <- inner_join(Total_Work_population, Unemploymentlang, by = "Country")

unemployment_percent_df <- merged_df["Country"]

newx_df <- inner_join(
  inner_join(GDP_df, Unemploymentlang, by = "Country"),
  unemployment_percent_df,
  by = "Country" 
)

breaks27 <- quantile(
  newx_df$GDP_growth_total,
  probs = c(0, 1/3, 2/3, 1),
  na.rm = TRUE
)

newxx_df <- newx_df %>%
  mutate(
    EconClass = cut(
      GDP_growth_total,
      breaks = breaks27,
      labels = c("Low growth", "Middle growth", "High growth"),
      include.lowest = TRUE
    )
  )

```

First, we used all available data to create a new column representing the total GDP growth. Based on this, we divided the countries into three equally sized groups. This allows us to compare unemployment rates relative to each country's level of economic growth

```{r show data}
head(newxx_df[, c(1,25)], n = 5)

```

We use this data to clearly show the unemployment rates in different types of countries with varying levels of growth, in order to identify a possible relationship between them. We use a boxplot for this, so the three different groups are clearly arranged side by side and easy to compare.

Variable 2: Cumulative GDP Growth

We also create variables to assess cumulative GDP growth before and after the COVID-19 pandemic.

```{r gen_var_beforre_covid}
GDP_df <- GDP_df %>%
  rowwise() %>%
  mutate(
    GDP_before_covid = round((prod(1 + c_across(`2015`:`2019`) / 100, na.rm = TRUE) - 1) * 100, 2))

```

```{r gen_var_after_covid}
GDP_df <- GDP_df %>%
  rowwise() %>%
  mutate(GDP_after_covid = round((prod(1 + c_across(`2020`:`2024`) / 100, na.rm = TRUE) - 1) * 100, 2)) %>%
  ungroup()
```

After creating these two variables, we created a new data frame called growth_df. In this data frame, we calculated the average GDP growth per year across all countries. We then used this data to create two plots: EU GDP Growth vs. Average Unemployment and Cumulative EU GDP Growth.

```{r gen_ data_frame_avg_GDP_growth}
gdp_growth <- GDP_df %>%
  select(all_of(as.character(2013:2024)))
avg_growth_per_year <- colMeans(gdp_growth, na.rm = TRUE) / 100 

growth_df <- data.frame(
  Year = 2013:2024,
  GrowthRate = avg_growth_per_year
)

growth_df <- growth_df %>%
  mutate(
    TotalGDP = cumprod(1 + GrowthRate) * 100
  )
```

We used the GDP_df data, which contains annual GDP growth rates from 2015 to 2024. Using c_across() and prod() inside mutate(), we calculate the compounded GDP growth rate for two periods:

-   Pre-COVID: 2015–2019
-   Post-COVID: 2020–2024

This transformation is informative because it helps assess how economic performance shifted due to the pandemic. By comparing these two variables, we can identify which countries experienced economic resilience and which struggled to recover.

```{r show data GDP}
head(GDP_df, 5)[ , c(1, 14, 15,16)]
head(growth_df)[,c(2,3)]
```

These variables provide a basis for analyzing and comparing GDP performance across countries before and after the pandemic. This allows us to assess the relative economic resilience and recovery of different European countries, and identify which economies grew more strongly or faced longer-term setbacks due to COVID-19.

We visualize the average cumulative GDP growth across EU countries from 2013 to 2024 in a line chart, setting the base year to 100. The red dashed line at 2020 highlights the start of the COVID-19 pandemic, allowing us to visually compare pre- and post-pandemic trends.

Variable 3:

To analyse

```{r gen_var_Growth_Rate}
gdp_growth <- GDP_df %>%
  select(all_of(as.character(2015:2024)))

avg_growth_per_year <- colMeans(gdp_growth, na.rm = TRUE) / 100

```

```{r gen_var_Total_GDP}
growth_df <- data.frame(
  Year = 2015:2024,
  GrowthRate = avg_growth_per_year
) %>%
  mutate(TotalGDP = cumprod(1 + GrowthRate) * 100)

unemp_df <- Unemploymentlang %>%
  summarise(across(all_of(as.character(2015:2024)), ~mean(as.numeric(.), na.rm = TRUE))) %>%
  pivot_longer(cols = everything(), names_to = "Year", values_to = "Unemployment") %>%
  mutate(Year = as.integer(Year))
```

## 3.3 Visualize temporal variation

```{r echo=FALSE}

ggplot(growth_df, aes(x = Year, y = TotalGDP)) +
  geom_line(color = "blue", size = 1) +
  geom_point(color = "blue") +
  geom_vline(xintercept = 2020, color = "red", linetype = "dashed", size = 1) +
  scale_x_continuous(breaks = 2013:2024) +  # elk jaar tonen
  labs(
    title = "Cumulative EU GDP Growth (2013–2024)",
    subtitle = "Red line marks the start of Covid-19 (2020), ",
    x = "Year",
    y = "Cumulative GDP Index"
  ) +
  theme_minimal()

```

## 3.4 Visualize spatial variation

```{r}
# Zorg dat kolomnamen kloppen
colnames(Total_Work_population)[1] <- "Country"
colnames(Unemploymentlang)[1] <- "Country"

# Merge datasets
merged_df <- inner_join(Total_Work_population, Unemploymentlang, by = "Country")

# Zet jaartallen
years <- 2015:2024
num_x <- paste0(years, ".x")
num_y <- paste0(years, ".y")

# Zorg dat kolommen numeriek zijn
merged_df[num_x] <- lapply(merged_df[num_x], as.numeric)
merged_df[num_y] <- lapply(merged_df[num_y], as.numeric)

# Bereken percentage werkloosheid per jaar
unemployment_percent_df <- merged_df %>%
  transmute(Country = Country)

for (year in years) {
  pop_col  <- paste0(year, ".x")
  unemp_col<- paste0(year, ".y")
  perc_col <- paste0(year, "_percent")
  
  unemployment_percent_df[[perc_col]] <-
    merged_df[[unemp_col]] / merged_df[[pop_col]] * 100
}

# Gemiddelde werkloosheid over de jaren
unemployment_percent_df$avg_unemp_percent <- round(
  rowMeans(unemployment_percent_df[ , grep("_percent$", names(unemployment_percent_df))], na.rm = TRUE),
  1
)

# Kaartdata koppelen
mapped_data_percent <- joinCountryData2Map(unemployment_percent_df,
                                           joinCode = "NAME",
                                           nameJoinColumn = "Country")

# Plot kaart
mapCountryData(
  mapped_data_percent,
  nameColumnToPlot = "avg_unemp_percent",
  catMethod = "quantiles",
  numCats = 3,
  mapTitle = "Average Unemployment in Europe (2015–2024)",
  colourPalette = c("green", "yellow", "red"),
  mapRegion = "Europe",
  addLegend = FALSE
)

legend(
  "topleft",
  legend = c("Low", "Medium", "High"),
  fill = c("green", "yellow", "red"),
  title = "Unemployment level",
  cex = 0.8
)
```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.5 Visualize sub-population variation

To explore how unemployment varies across different levels of economic performance, we classified European countries into three economic classes: Low growth, Middle growth, and High growth. These classes were derived by calculating the total cumulative GDP growth for each country between 2015 and 2024. This was done using annual GDP growth data retrieved from Eurostat (2025), and countries were then grouped into tertiles based on their total growth performance.

We then compared the unemployment rates in 2024 across these economic classes. The unemployment data was merged with the GDP dataset to form a unified data frame, which enabled us to conduct this sub-population analysis.

```{r visualise_subpopulations, echo=FALSE}
# Bereken totale GDP-groei per land
GDP_df <- GDP_df %>%
  rowwise() %>%
  mutate(
    GDP_growth_total = round((prod(1 + c_across(`2015`:`2024`) / 100, na.rm = TRUE) - 1) * 100, 2)
  ) %>%
  ungroup()

colnames(GDP_df)[1] <- "Country"
colnames(Unemploymentlang)[1] <- "Country"

# Combineer datasets
newx_df <- inner_join(
  inner_join(GDP_df, Unemploymentlang, by = "Country"),
  unemployment_percent_df,
  by = "Country" 
)

# Verdeel landen in drie groeigroepen op basis van GDP-groei
breaks27 <- quantile(
  newx_df$GDP_growth_total,
  probs = c(0, 1/3, 2/3, 1),
  na.rm = TRUE
)

newxx_df <- newx_df %>%
  mutate(
    GDP_Growth_Group = cut(
      GDP_growth_total,
      breaks = breaks27,
      labels = c("Low GDP Growth", "Medium GDP Growth", "High GDP Growth"),
      include.lowest = TRUE
    )
  )

# Genereer boxplot
ggplot(newxx_df, aes(x = GDP_Growth_Group, y = `2024_percent`, fill = GDP_Growth_Group)) +
  geom_boxplot() +
  labs(
    title = "Unemployment Rate in 2024 by GDP Growth",
    x = "GDP Growth Group (2015–2024)",
    y = "Unemployment Rate (%) in 2024",
    fill = "GDP Growth Group"
  ) +
  theme_minimal()


```

The boxplot reveals that countries in the High growth category tend to have lower unemployment rates, with both a lower median and smaller spread. In contrast, Low growth and Middle growth groups show greater variability and higher central unemployment values. These patterns suggest that stronger economic performance over the past decade may be associated with improved labor market outcomes. However, differences within each group also highlight the role of country-specific factors beyond GDP growth alone.

This sub-population comparison illustrates the unequal impact of unemployment across economic contexts, reinforcing the broader relationship between macroeconomic growth and labor market performance in Europe.

## 3.6 Event analysis

In this analyses we analyse 2 key veriables, we compare avarage GDP growth over the years in the EU, and we compare this to the avarage unemployment rate. We do this by adding two y-axes, on the left side we have cumulative GDP index, with its base year in 2015 we start with 100%.

```{r analysis, echo=FALSE}

#combine datasets
combo_df <- left_join(growth_df, unemp_df, by = "Year")

#Build the plot

ggplot(combo_df, aes(x = Year)) +
  geom_line(aes(y = TotalGDP), color = "#0072B2", size = 1.3) +
  geom_point(aes(y = TotalGDP), color = "#0072B2", size = 2) +
  
  geom_line(aes(y = Unemployment / 7.5), color = "#E69F00", size = 1.3) +
  geom_point(aes(y = Unemployment / 7.5), color = "#E69F00", size = 2) +
  
  scale_y_continuous(
    limits = c(0, 135),
    name = "Cumulative GDP Index (Base year = 100)",
    sec.axis = sec_axis(~.*7.5, name = "Average Unemployment (in thousands)", breaks = seq(0, 1000, 250))
  ) +
  
  scale_x_continuous(breaks = 2015:2024) +
  
  labs(
    title = "EU GDP Growth vs. Average Unemployment (2015–2024)",
    subtitle = "Left: GDP Index (blue), Right: Avg. Unemployment (orange)",
    x = "Year"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.title.y.left = element_text(color = "#0072B2", face = "bold"),
    axis.title.y.right = element_text(color = "#E69F00", face = "bold"),
    axis.text.y.left = element_text(color = "#0072B2"),
    axis.text.y.right = element_text(color = "#E69F00")
  )
```

Here you provide a description of why the plot above is relevant to your specific social problem.

# Part 4 - Discussion

## 4.1 Discuss your findings

# Part 5 - Reproducibility

## 5.1 Github repository link

Provide the link to your PUBLIC repository here: ...

## 5.2 Reference list

Eurostat. (n.d.). Unemployment by sex and age – monthly data [Data set]. European Commission.

<https://ec.europa.eu/eurostat/databrowser/view/lfsa_ugad/default/table?lang=en>

Eurostat. (18/06/2025). Real GDP growth rate - volume (tec00115). European Commission.

<https://ec.europa.eu/eurostat/databrowser/view/tec00115/default/table?lang=en>

Eurostat. (12/06/2025). Population, aged 15-74 - EU labour force survey (tipslm16).

<https://ec.europa.eu/eurostat/databrowser/view/tipslm16/default/table?lang=en>
