---
title: "Research Unemployment to GDP"
author: "Studentnames and studentnumbers here"
date: "`r Sys.Date()`"
output: pdf_document
---

# Set-up your environment

```{r package_install, include=FALSE}
#install.packages("tidyverse")
#install.packages("readxl")
#install.packages("rworldmap")
#install.packages("dplyr") 
#install.packages("ggplot2")

```

```{r packages}
require(tidyverse)
require(readxl)
require(rworldmap)
require(dplyr)
require(ggplot2)

```

# Title Page

Chris van Weerdenburg, Arsema Tewelde, Alan Peralta Camacho, Benjamin Mutlu, Cihan Öztürk, Israa Shidai, Jeremey Li

Tutorial group number: 3

Tutorial lecturer: Jack Fitzgerald

# Part 1 - Identify a Social Problem

Use APA referencing throughout your document.

## 1.1 Describe the Social Problem

Include the following:

-   Why is this relevant?

-   ...

# Part 2 - Data Sourcing

We have the following datasets:

-   **Unemployment by sex, age and duration of unemployment (Eurostat)**

-   **Real GDP growth rate - volume (Eurostat)**

These datasets are published and sourced by Eurostat. Eurostat is the statistical office of the European Union. They source their data through the ESS which is the European Statistical System. These datasets are also collected by each country's own National Statistical Institute for example the data Eurostat collects from the Netherlands is from the CBS (Centraal Bureau Statistiek). This data collection is done according to established EU regulations and standards, ensuring a degree of harmonization across the EU. The national authorities then transmit the verified and analyzed data to Eurostat.

These datasets have some differences but also similarities.

**Key differences:**

-   GDP per capita relates to overall economic performances, while youth unemployment reflects specific labor market outcomes.

-   GDP includes the whole population; youth unemployment only concerns young people who are economically active.

-   GDP is measured in economic value (adjusted for PPS); youth unemployment is a rate (percentage).

**Complementary:**

-   A country may have high GDP per capita but also high youth unemployment, indicating that economic gains are not evenly distributed or accessible.

-   Using both datasets together provides a more comprehensive view of socio-economic well-being, economic performance (GDP) and inclusiveness or opportunity (youth employment).

-   This combination can be especially useful for analyzing whether economic growth is benefiting younger generations.

These two datasets were a perfect match for our topic because they gave us an in depth look into the correlation between the GDP growth rate and Unemployment rates but also the unemployment rates and sex, age and duration of unemployment these datasets were crucial for us to conclude whether Covid-19 had an impact on unemployment rates.

The datasets we used were quite complete however some countries the data was not sourced for example: Bosnia and Herzegovina and Lichtenstein. However we did add in a new data set because the other data sets did not give as a good insight about the percentage of the population being unemployed. It just gave us the amount of people that are / were unemployed. This was a slight hiccup since for example Germany has a bigger population than Norway. Other than that were there no complaints about the datasets, and were they very explicit.

## 2.1 Load in the data

```{r loading_data, include=FALSE}
GDP_df <- read_xlsx(path = "tec00115_page_spreadsheet.xlsx", "Sheet 1") 
Unemploymentlang <- read_xlsx(path = "lfsa_ugad$defaultview_spreadsheet.xlsx", "Sheet 1")
unemp_men <- readxl::read_xlsx("lfsa_ugad$defaultview_spreadsheet.xlsx", sheet = "Sheet 6")
unemp_women <- readxl::read_xlsx("lfsa_ugad$defaultview_spreadsheet.xlsx", sheet = "Sheet 11")

```

midwest is an example dataset included in the tidyverse package

## 2.2 Provide a short summary of the dataset(s)

```{r echo=TRUE}
head(GDP_df[,(1:3)], n = 5)
head(Unemploymentlang[,(1:3)], n = 5)
```

In this case we see 28 variables, but we miss some information on what units they are in. We also don't know anything about the year/moment in which this data has been captured.

These are things that are usually included in the metadata of the dataset. For your project, you need to provide us with the information from your metadata that we need to understand your dataset of choice.

## 2.3 Describe the type of variables included

Think of things like:

-   Do the variables contain health information or SES information?

-   Have they been measured by interviewing individuals or is the data coming from administrative sources?

*For the sake of this example, I will continue with the assignment...*

# Part 3 - Quantifying

## 3.1 Data cleaning

We will start with the unemployment dataset. We will remove the first two columns, as they are not relevant for our analysis.

```{r data_cleaning_Unemployment, echo=FALSE}

Unemploymentlang <- Unemploymentlang[14:49, ]
Unemploymentlang <- Unemploymentlang[, -c( 3, 5, 7 ,9 ,11 ,13 ,15 ,17 ,19 ,21)]
rownames(Unemploymentlang) <- NULL
colnames(Unemploymentlang)[2:11] <- as.character(2015:2024)

```

```{r show_data, echo=TRUE}
head(Unemploymentlang, n = 3)
```

```{r data_cleaning_GDP, include=FALSE}
GDP_df <- GDP_df[10:52, ]
rownames(GDP_df) <- NULL
GDP_df <- GDP_df[, colSums(!is.na(GDP_df)) > 0]
GDP_df <- GDP_df[, -c( 5, 12, 14, 16, 18)]
colnames(GDP_df)[2:13] <- as.character(2013:2024)
GDP_df[ , as.character(2013:2024)] <- lapply(GDP_df[ , as.character(2013:2024)], as.numeric)
GDP_df <- GDP_df[-c(1,2,3),]
```

```{r show_data_GDP, echo=TRUE}
head(GDP_df, n = 3)

```

```{r data_cleaning_unemp_men, include=FALSE}
unemp_men <- unemp_men[14:49, ]
unemp_men <- unemp_men[, -c( 3, 5, 7 ,9 ,11 ,13 ,15 ,17 ,19 ,21)]
rownames(unemp_men) <- NULL
colnames(unemp_men)[2:11] <- as.character(2015:2024)
unemp_men[, 2:11] <- lapply(unemp_men[, 2:11], function(x) as.numeric(as.character(x)))

```

```{r show data unemp_men}
head(unemp_men, n = 3)
```

```{r data_cleaning_unemp_women, include=FALSE}
unemp_women <- unemp_women[14:49, ]
unemp_women <- unemp_women[, -c( 3, 5, 7 ,9 ,11 ,13 ,15 ,17 ,19 ,21)]
rownames(unemp_women) <- NULL
colnames(unemp_women)[2:11] <- as.character(2015:2024)
unemp_women[, 2:11] <- lapply(unemp_women[, 2:11], function(x) as.numeric(as.character(x)))
```

```{r show data unemp_women}
head(unemp_women, n = 3)
```

Please use a separate 'R block' of code for each type of cleaning. So, e.g. one for missing values, a new one for removing unnecessary variables etc.

## 3.2 Generate necessary variables

In this section, we generate and explain two new variables that are useful for our analysis of unemployment and GDP growth trends in Europe.

Variable 1: Average Unemployment by Gender

To compare long-term unemployment trends across genders and countries, we calculate the average unemployment rate for men and women over a 10-year period.

```{r gen_var1}
unemp_men$avg_men <- rowMeans(unemp_men[, 2:11], na.rm = TRUE)
unemp_women$avg_women <- rowMeans(unemp_women[, 2:11], na.rm = TRUE)

Unemploymentlang$avg_women <- unemp_women$avg_women
Unemploymentlang$avg_men <- unemp_men$avg_men

```

We used the unemp_men and unemp_women data frames, which contain annual unemployment rates from 2013 to 2022, sourced from Eurostat (2024). The rowMeans() function computes the average rate per country. The result is stored in the Unemploymentlang data frame for easier visualization.

These new variables are useful because they allow us to summarize a decade of data into a single value per gender, enabling straightforward comparison across countries. This helps highlight gender disparities and identify which European countries have the largest gaps in unemployment.

```{r show data}
head(Unemploymentlang[, c(1, 12, 13)], n = 5)

```

We will use these variables to create a geographical plot that visualizes average unemployment rates across European countries from 2015 to 2024, using color coding to indicate the severity of unemployment. This visualization will help us analyze regional unemployment patterns and identify countries with relatively higher or lower unemployment, as well as explore potential gender-based disparities.

Variable 2: Cumulative GDP Growth

We also create variables to assess cumulative GDP growth before and after the COVID-19 pandemic.

```{r gen_var2}
GDP_df <- GDP_df %>%
  rowwise() %>%
  mutate(
    GDP_before_covid = round((prod(1 + c_across(`2015`:`2019`) / 100, na.rm = TRUE) - 1) * 100, 2),
    GDP_after_covid  = round((prod(1 + c_across(`2020`:`2024`) / 100, na.rm = TRUE) - 1) * 100, 2)
  ) %>%
  ungroup()

gdp_growth <- GDP_df %>%
  select(all_of(as.character(2013:2024)))
avg_growth_per_year <- colMeans(gdp_growth, na.rm = TRUE) / 100  # omzetting naar factor
growth_df <- data.frame(
  Year = 2013:2024,
  GrowthRate = avg_growth_per_year
)

growth_df <- growth_df %>%
  mutate(
    TotalGDP = cumprod(1 + GrowthRate) * 100  # basisjaar = 100
  )

```

We used the GDP_df data, which contains annual GDP growth rates from 2015 to 2024. Using c_across() and prod() inside mutate(), we calculate the compounded GDP growth rate for two periods:

-   Pre-COVID: 2015–2019
-   Post-COVID: 2020–2024

This transformation is informative because it helps assess how economic performance shifted due to the pandemic. By comparing these two variables, we can identify which countries experienced economic resilience and which struggled to recover.

```{r show data GDP}
head(GDP_df, 5)[ , c(1, 14, 15)]
```

These variables provide a basis for analyzing and comparing GDP performance across countries before and after the pandemic. This allows us to assess the relative economic resilience and recovery of different European countries, and identify which economies grew more strongly or faced longer-term setbacks due to COVID-19.

We visualize the average cumulative GDP growth across EU countries from 2013 to 2024 in a line chart, setting the base year to 100. The red dashed line at 2020 highlights the start of the COVID-19 pandemic, allowing us to visually compare pre- and post-pandemic trends.

## 3.3 Visualize temporal variation

```{r echo=FALSE}

ggplot(growth_df, aes(x = Year, y = TotalGDP)) +
  geom_line(color = "blue", size = 1) +
  geom_point(color = "blue") +
  geom_vline(xintercept = 2020, color = "red", linetype = "dashed", size = 1) +
  scale_x_continuous(breaks = 2013:2024) +  # elk jaar tonen
  labs(
    title = "Cumulative EU GDP Growth (2013–2024)",
    subtitle = "Base year = 100, ",
    x = "Year",
    y = "Cumulative GDP Index"
  ) +
  theme_minimal()

```

## 3.4 Visualize spatial variation

```{r visualise_map, echo=FALSE}
              
 # Bereken gemiddelde werkloosheid per land (2015–2024)
  unemp_map <- Unemploymentlang %>%
  mutate(across(`2015`:`2024`, ~as.numeric(as.character(.)))) %>%
  mutate(avg_unemp = rowMeans(select(., `2015`:`2024`), na.rm = TRUE)) %>%
  select(Country = 1, avg_unemp)
  
  # Corrigeer landnamen
  unemp_map$Country[unemp_map$Country == "Czechia"] <- "Czech Rep."
  unemp_map$Country[unemp_map$Country == "Bosnia and Herzegovina"] <- "Bosnia and Herz."
  unemp_map$Country[unemp_map$Country == "North Macedonia"] <- "Macedonia"
  unemp_map$Country[unemp_map$Country == "Türkiye"] <- "Turkey"
               
 # Koppel de data aan de wereldkaart
  mapped_data <- joinCountryData2Map(unemp_map,
                                     joinCode = "NAME",
                                     nameJoinColumn = "Country")
               
 # Maak de kaart met focus op Europa
  mapCountryData(mapped_data,
                 nameColumnToPlot = "avg_unemp",
                 catMethod = "quantiles",
                 numCats = 3,
                 mapTitle = "Average Unemployment in Europe (2015–2024)\nGreen = Low, Yellow = Medium, Red = High",
                 colourPalette = c("green", "yellow", "red"),
                 mapRegion = "Europe")
                
  
               
```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.5 Visualize sub-population variation

What is the poverty rate by state?

```{r visualise_subpopulations, echo=FALSE}
# Pivot both datasets to long form and add a Gender column
men_long <- unemp_men %>%
  select(Country = 1, `2015`:`2024`) %>%
  pivot_longer(`2015`:`2024`, names_to = "Year", values_to = "Unemployment") %>%
  mutate(Gender = "Men")

women_long <- unemp_women %>%
  select(Country = 1, `2015`:`2024`) %>%
  pivot_longer(`2015`:`2024`, names_to = "Year", values_to = "Unemployment") %>%
  mutate(Gender = "Women")

# Combine and factor Year
unemp_gender <- bind_rows(men_long, women_long) %>%
  mutate(Year = factor(Year, levels = as.character(2015:2024)))

# Plot the boxplot with a shorter y–axis
ggplot(unemp_gender, aes(x = Year, y = Unemployment, fill = Gender)) +
  geom_boxplot(
    position      = position_dodge(width = 0.8),
    outlier.shape = NA
  ) +
  scale_y_continuous(
    limits = c(0, 30),         # set the lower & upper bounds
    expand = expansion(0, 0)   # remove padding at axis ends
  ) +
  labs(
    title = "Unemployment Rates: Men vs Women (EU Countries, 2015–2024)",
    x     = "Year",
    y     = "Unemployment (%)",
    fill  = "Gender"
  ) +
  theme_minimal()
```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.6 Event analysis

Analyze the relationship between two variables.

```{r analysis, echo=FALSE}
# ///Temporal Variation Plot/EU Economic Development: GDP Index and Average Unemployment (2015–2024)///
# 📊 1. Maak je samengestelde GDP-index klaar (2015–2024)
gdp_growth <- GDP_df %>%
  select(all_of(as.character(2015:2024)))

avg_growth_per_year <- colMeans(gdp_growth, na.rm = TRUE) / 100

growth_df <- data.frame(
  Year = 2015:2024,
  GrowthRate = avg_growth_per_year
) %>%
  mutate(TotalGDP = cumprod(1 + GrowthRate) * 100)

# 📊 2. Maak gemiddelde werkloosheid klaar (2015–2024)
unemp_df <- Unemploymentlang %>%
  summarise(across(all_of(as.character(2015:2024)), ~mean(as.numeric(.), na.rm = TRUE))) %>%
  pivot_longer(cols = everything(), names_to = "Year", values_to = "Unemployment") %>%
  mutate(Year = as.integer(Year))

# 🔗 3. Combineer de twee datasets
combo_df <- left_join(growth_df, unemp_df, by = "Year")

# 🎨 4. Plot met dubbele y-as en mooie styling

ggplot(combo_df, aes(x = Year)) +
  # Blauwe GDP-lijn (links)
  geom_line(aes(y = TotalGDP), color = "#0072B2", size = 1.3) +
  geom_point(aes(y = TotalGDP), color = "#0072B2", size = 2) +
  
  # Oranje werkloosheidslijn (rechts, herschaald)
  geom_line(aes(y = Unemployment / 7.5), color = "#E69F00", size = 1.3) +
  geom_point(aes(y = Unemployment / 7.5), color = "#E69F00", size = 2) +
  
  # Y-assen
  scale_y_continuous(
    limits = c(0, 135),
    name = "Cumulative GDP Index (Base year = 100)",
    sec.axis = sec_axis(~.*7.5, name = "Average Unemployment (in thousands)", breaks = seq(0, 1000, 250))
  ) +
  
  # X-as per jaar tonen
  scale_x_continuous(breaks = 2015:2024) +
  
  # Titel & opmaak
  labs(
    title = "EU GDP Growth vs. Average Unemployment (2015–2024)",
    subtitle = "Left: GDP Index (blue), Right: Avg. Unemployment (orange)",
    x = "Year"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.title.y.left = element_text(color = "#0072B2", face = "bold"),
    axis.title.y.right = element_text(color = "#E69F00", face = "bold"),
    axis.text.y.left = element_text(color = "#0072B2"),
    axis.text.y.right = element_text(color = "#E69F00")
  )
```

Here you provide a description of why the plot above is relevant to your specific social problem.

# Part 4 - Discussion

## 4.1 Discuss your findings

# Part 5 - Reproducibility

## 5.1 Github repository link

Provide the link to your PUBLIC repository here: ...

## 5.2 Reference list

Use APA referencing throughout your document.

Eurostat. (n.d.). Unemployment by sex and age – monthly data [Data set]. European Commission. <https://ec.europa.eu/eurostat/databrowser/view/lfsa_ugad/default/table?lang=en>

Eurostat. (2024). Real GDP growth rate - volume (tec00115). European Commission. <https://ec.europa.eu/eurostat/databrowser/view/tec00115/default/table?lang=en>
