---
title: "Research Unemployment to GDP"
author: "Studentnames and studentnumbers here"
date: "`r Sys.Date()`"
output: pdf_document
---

# Set-up your environment

```{r package_install, include=FALSE}
install.packages("tidyverse")
install.packages("readxl")
install.packages("rworldmap")
install.packages("dplyr") 
install.packages("ggplot2")

```

```{r packages, include=FALSE}
require(tidyverse)
require(readxl)
require(rworldmap)
require(dplyr)
require(ggplot2)

```



# Title Page

Chris van Weerdenburg, Arsema Tewelde, Alan Peralta Camacho, Benjamin Mutlu, Cihan Öztürk, Israa Shidai, Jeremey Li

Include the tutorial group number

Include your tutorial lecturer's name

# Part 1 - Identify a Social Problem

Use APA referencing throughout your document.

## 1.1 Describe the Social Problem

Include the following:

-   Why is this relevant?

-   ...

# Part 2 - Data Sourcing

## 2.1 Load in the data

```{r loading_data, include=FALSE}
GDP_df <- read_xlsx(path = "tec00115_page_spreadsheet.xlsx", "Sheet 1") 
Unemploymentlang <- read_xlsx(path = "lfsa_ugad$defaultview_spreadsheet.xlsx", "Sheet 1")
unemp_men <- readxl::read_xlsx("lfsa_ugad$defaultview_spreadsheet.xlsx", sheet = "Sheet 6")
unemp_women <- readxl::read_xlsx("lfsa_ugad$defaultview_spreadsheet.xlsx", sheet = "Sheet 11")

```

midwest is an example dataset included in the tidyverse package

## 2.2 Provide a short summary of the dataset(s)

```{r echo=TRUE}
head(GDP_df[, 1:10], n = 20)
head(Unemploymentlang[, 1:10], n = 20)
head(unemp_men[, 1:10], n = 20)
head(unemp_women[, 1:10], n = 20)
```

In this case we see 28 variables, but we miss some information on what units they are in. We also don't know anything about the year/moment in which this data has been captured.

These are things that are usually included in the metadata of the dataset. For your project, you need to provide us with the information from your metadata that we need to understand your dataset of choice.


## 2.3 Describe the type of variables included

Think of things like:

-   Do the variables contain health information or SES information?

-   Have they been measured by interviewing individuals or is the data coming from administrative sources?

*For the sake of this example, I will continue with the assignment...*



# Part 3 - Quantifying

## 3.1 Data cleaning
We will start with the unemployment dataset. We will remove the first two columns, as they are not relevant for our analysis.

```{r data_cleaning_Unemployment, echo=FALSE}

Unemploymentlang <- Unemploymentlang[14:49, ]
Unemploymentlang <- Unemploymentlang[, -c( 3, 5, 7 ,9 ,11 ,13 ,15 ,17 ,19 ,21)]
rownames(Unemploymentlang) <- NULL
colnames(Unemploymentlang)[2:11] <- as.character(2015:2024)

```

```{r show_data, echo=TRUE}
head(Unemploymentlang, n = 20)
```


```{r data_cleaning_GDP, include=FALSE}
GDP_df <- GDP_df[10:52, ]
rownames(GDP_df) <- NULL
GDP_df <- GDP_df[, colSums(!is.na(GDP_df)) > 0]
GDP_df <- GDP_df[, -c( 5, 12, 14, 16, 18)]
colnames(GDP_df)[2:13] <- as.character(2013:2024)
GDP_df[ , as.character(2013:2024)] <- lapply(GDP_df[ , as.character(2013:2024)], as.numeric)
GDP_df <- GDP_df[-c(1,2,3),]
```

```{r show_data_GDP, echo=TRUE}
head(GDP_df, n = 20)

```
```{r data_cleaning_unemp_men, include=FALSE}
unemp_men <- unemp_men[14:49, ]
unemp_men <- unemp_men[, -c( 3, 5, 7 ,9 ,11 ,13 ,15 ,17 ,19 ,21)]
rownames(unemp_men) <- NULL
colnames(unemp_men)[2:11] <- as.character(2015:2024)
unemp_men[, 2:11] <- lapply(unemp_men[, 2:11], function(x) as.numeric(as.character(x)))

```

```{r show data unemp_men}
head(unemp_men, n = 20)
```

```{r data_cleaning_unemp_women, include=FALSE}
unemp_women <- unemp_women[14:49, ]
unemp_women <- unemp_women[, -c( 3, 5, 7 ,9 ,11 ,13 ,15 ,17 ,19 ,21)]
rownames(unemp_women) <- NULL
colnames(unemp_women)[2:11] <- as.character(2015:2024)
unemp_women[, 2:11] <- lapply(unemp_women[, 2:11], function(x) as.numeric(as.character(x)))
```


```{r show data unemp_women}
head(unemp_women, n = 20)
```


Please use a separate 'R block' of code for each type of cleaning. So, e.g. one for missing values, a new one for removing unnecessary variables etc.

## 3.2 Generate necessary variables

Variable 1
```{r gen_var1}
unemp_men$avg_men <- rowMeans(unemp_men[, 2:11], na.rm = TRUE)
unemp_women$avg_women <- rowMeans(unemp_women[, 2:11], na.rm = TRUE)

Unemploymentlang$avg_women <- unemp_women$avg_women
Unemploymentlang$avg_men <- unemp_men$avg_men

```

```{r show data}
head(Unemploymentlang[, c(1, 12, 13)], n = 10)

```

Variable 2
```{r gen_var2}
GDP_df <- GDP_df %>%
  rowwise() %>%
  mutate(
    GDP_before_covid = round((prod(1 + c_across(`2015`:`2019`) / 100, na.rm = TRUE) - 1) * 100, 2),
    GDP_after_covid  = round((prod(1 + c_across(`2020`:`2024`) / 100, na.rm = TRUE) - 1) * 100, 2)
  ) %>%
  ungroup()
```

```{r show data GDP}

head(GDP_df, 10)[ , c(1, 14, 15)]
```

## 3.3 Visualize temporal variation

```{r}

```

## 3.4 Visualize spatial variation

```{r visualise_map}

```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.5 Visualize sub-population variation

What is the poverty rate by state?

```{r visualise_subpopulations}
dataset$inmetro <- dataset$inmetro %>% as.factor()
# Boxplot of poverty rate by state using the 'midwest' dataset
ggplot(dataset, aes(x = inmetro, y = percadultpoverty)) +
  geom_boxplot() +
  labs(
    title = "Distribution of Poverty Rates by Metropolitan status (Midwest counties)",
    x = "Metropolitan Area",
    y = "Poverty Rate of Adults (%)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right"
  )
```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.6 Event analysis

Analyze the relationship between two variables.

```{r analysis}

```

Here you provide a description of why the plot above is relevant to your specific social problem.

# Part 4 - Discussion

## 4.1 Discuss your findings

# Part 5 - Reproducibility

## 5.1 Github repository link

Provide the link to your PUBLIC repository here: ...

## 5.2 Reference list

Use APA referencing throughout your document.
